---
title: "Class 11: Extra Credit"
author: "Aparajita Pranjal"
format: pdf
editor: visual
---

**Q13:** Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

Reading in file:

```{r}
data_new <- read.table("datafile.txt")
head(data_new)
```

Counting sample size of each genotype:

```{r}
table(data_new$geno)
```

Obtaining median expression levels of each genotype:

```{r}
AA <- subset(data_new, geno=="A/A")
AG <- subset(data_new, geno=="A/G" | geno=="G/A")
GG <- subset(data_new, geno=="G/G")

median(AA$exp)
median(AG$exp)
median(GG$exp)
```

**Q14:** Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Generating boxplot using ggplot:

```{r}
library(ggplot2)

p <- ggplot(data_new) +
  aes(data_new$geno, data_new$exp, fill = data_new$geno) +
  geom_boxplot()

p + ggtitle("Expression of ORMDL3 across 3 genotypes") + 
    xlab("Genotype") +
    ylab("Expression Level") +
    theme(legend.position="none")
```

**The SNP does affect the expression level of ORMDL3 as the median value for A/A is 31.2 and G/G is 20.1. A difference of about 11 is significant for gene expression.**
