---
title: "Class 8: ML Mini Project"
author: "Aparajita Pranjal"
format: pdf
editor: visual
---

## Exploratory Data Analysis

The .csv file was downloaded and read in as a new variable.

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
# Create diagnosis vector for later 
diagnosis <- factor(wisc.df[,1])
```

-   **Q1**. How many observations are in this dataset? **569**

    ```{r}
    dim(wisc.data)
    ```

-   **Q2**. How many of the observations have a malignant diagnosis? **212**

    ```{r}
    table(diagnosis, exclude = "B")

    ```

-   **Q3**. How many variables/features in the data are suffixed with `_mean`? **10 variables**

```{r}
col_names <- colnames(wisc.data)
length(grep("_mean", col_names))
```

## PCA Analysis

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)

# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE)

# Look at summary of results
summary(wisc.pr)
```

-   **Q4**. From your results, what proportion of the original variance is captured by the first principal components (PC1)? **0.4427**

-   **Q5**. How many principal components (PCs) are required to describe at least 70% of the original variance in the data? **3 components**

-   **Q6**. How many principal components (PCs) are required to describe at least 90% of the original variance in the data? **7 components**

# Interpreting PCA Results

```{r}
biplot(wisc.pr)
```

-   **Q7.** What stands out to you about this plot? Is it easy or difficult to understand? Why?

    **The function plots all the points on the graph and labels them which makes it difficult to read and understand the different groups/trends.**

```{r}
plot(wisc.pr$x , col = diagnosis, 
     xlab = "PC1", ylab = "PC2")
```

-   **Q8.** Generate a similar plot for principal components 1 and 3. What do you notice about these plots? **They have a larger range on the y-axis for PCA3 as the data points are a bit more spread out.**

```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[,1],wisc.pr$x[,3], col = diagnosis, 
     xlab = "PC1", ylab = "PC3")
```

# Graphing with ggplot

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col= diagnosis) + 
  geom_point()
```

## Variance Explained

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)

# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

Scree Plot

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

-   **Q9.** For the first principal component, what is the component of the loading vector (i.e. `wisc.pr$rotation[,1]`) for the feature `concave.points_mean`? This tells us how much this original feature contributes to the first PC. **-0.26085376**

```{r}
wisc.pr$rotation[,1]
```

# Hierarchical Clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

data.dist <- dist(data.scaled)

wisc.hclust <- hclust(data.dist, method = "complete")
```

-   **Q10.** Using the `plot()` and `abline()` functions, what is the height at which the clustering model has 4 clusters? **Height is 19**

```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2, )
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h = 19)
table(wisc.hclust.clusters, diagnosis)
```

-   **Q12.** Which method gives your favorite results for the same `data.dist` dataset? Explain your reasoning.

    **Ward.D2 gives the best results for the same dataset as it makes visualizing the differences between clusters much easier as it squares the differences whereas complete takes in all the data points and creates a large dendogram with tighter clusters as we go further down the tree whereas ward.D\@ distinguishes them higher up.**

Clustering PCA Results

```{r}
wisc.pr.hclust <- hclust(data.dist, method = "ward.D2")
plot(wisc.pr.hclust)

grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
table(grps, diagnosis)

plot(wisc.pr$x[,1:2], col=grps)

plot(wisc.pr$x[,1:2], col=diagnosis)

g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)

# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)

```

New Plot

```{r}
#install.packages("rgl")
#library(rgl)
#plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
```

Clustering Continued

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)

table(wisc.pr.hclust.clusters , diagnosis)

table(wisc.hclust.clusters, diagnosis)
```

-   **Q13**. How well does the newly created model with four clusters separate out the two diagnoses? **The new model further sub-categorizes the two diagnoses providing a more detailed breakdown of benign versus malignant.**

-   **Q14**. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the `table()` function to compare the output of each model (`wisc.km$cluster` and `wisc.hclust.clusters`) with the vector containing the actual diagnoses. **Before PCA the clustering does not provide information on outliers or potential false readings which after PCA there is a difference if a few individuals that was separated which is more useful in obtaining specific details from the dataset and not just looking it as it overall.**

# Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc

plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

-   **Q16.** Which of these new patients should we prioritize for follow up based on your results? **Group 2 as they belong to the malignant category and should therefore be brought in for further testing before the benign patients.**
