---
title: "Class 17: Vaccine Mini Project"
author: "Aparajita Pranjal"
date: "2023-05-31"
format: pdf
editor: visual
---

## Importing data

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

-   **Q1.** What column details the total number of people fully vaccinated? **Column 10 - "persons fully vaccinated"**

    ```{r}
    colnames(vax)
    ```

-   **Q2.** What column details the Zip code tabulation area?

    **Column 2 - "zip_code_tabulation_area"**

-   **Q3.** What is the earliest date in this dataset?

    **2021-01-05**

-   **Q4.** What is the latest date in this dataset?

    **2023-05-23**

Summary of data:

```{r}
skimr::skim_without_charts(vax)
```

-   **Q5.** How many numeric columns are in this dataset?

    **14 numeric columns**

-   **Q6.** Note that there are \"missing values\" in the dataset. How many `NA` values there in the `persons_fully_vaccinated` column? **17711**

    ```{r}
    sum( is.na(vax$persons_fully_vaccinated) )
    ```

-   **Q7.** What percent of `persons_fully_vaccinated` values are missing (to 2 significant figures)?

    **(17711/220500) \* 100 = 8.03%**

-   **Q8.** \[Optional\]: Why might this data be missing?

    **If they were vaccinated in another state their medical records may not have been updated.**

Using lubridate:

```{r}
library(lubridate)
```

```{r}
today()
```

Incorrect use of package:

```{r}
# This will give an Error!
#today() - vax$as_of_date[1]
```

Changing the date into the correct format:

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Difference from today and first date of reported vaccine:

```{r}
today() - vax$as_of_date[1]
```

To find the difference between the first and last reported date:

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

-   **Q9.** How many days have passed since the last update of the dataset? **8 days**

    ```{r}
    today() - ymd(vax$as_of_date[nrow(vax)])
    ```

-   **Q10.** How many unique dates are in the dataset (i.e. how many different dates are detailed)? **125 unique dates**

```{r}
length(unique(vax$as_of_date))
```

## Working with zipcodes:

```{r}
library(zipcodeR)
```

Coordinates of La Jolla zipcode:

```{r}
geocode_zip('92037')
```

Calculating the distance between zipcode:

```{r}
zip_distance('92037','92109')
```

For census data:

```{r}
reverse_zipcode(c('92037', "92109") )
```

Census data for vaccine file zipcodes:

```{r}
# Pull data for all ZIP codes in the dataset
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

Focusing on San Diego using base R:

```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county == "San Diego", ]
dim(sd)
```

Using dyplr:

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Filtering data further:

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

-   **Q11.** How many distinct zip codes are listed for San Diego County? **107 unique zip codes**

    ```{r}
    length(unique(sd$zip_code_tabulation_area))
    ```

-   **Q12.** What San Diego County Zip code area has the largest population in this dataset? **92154**

    ```{r}
    which.max(sd.10$tot_population)
    sd.10$zip_code_tabulation_area[64]
    ```

```{=html}
<!-- -->
```
-   **Q13.** What is the overall average (with 2 decimal numbers) \"Percent of Population Fully Vaccinated\" value for all San Diego \"County\" as of \"2023-05-23\"? **0.74**

    ```{r}
    sd.23 <- filter(sd, as_of_date =='2023-05-23')
    sd.23

    mean(sd.23$percent_of_population_fully_vaccinated, na.rm = T)
    ```

-   **Q14.** Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of \"2023-05-23\"?

```{r}
library(ggplot2)

p <- ggplot(sd.23) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(binwidth = 0.1)

p + ggtitle("Histogram of Vaccination Rate Across San Diego County") +
  xlab("Percent of Population Fully Vaccinated in a Zipcode Area") + ylab("Count (Zipcode Areas)")

```

Focusing on UCSD:

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

-   **Q15**. Using **ggplot** make a graph of the vaccination rate time course for the 92037 ZIP code area:

    ```{r}
    p <- ggplot(ucsd) +
      aes(as_of_date,
          percent_of_population_fully_vaccinated) +
      geom_point() +
      geom_line(group=1) +
      ylim(c(0,1)) +
      labs(x = "Date", y="Percent Vaccinated")
    p + ggtitle("Vaccination Rate for La Jolla, CA 92037")
    ```

Comparing similar sized areas:

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-05-23")

head(vax.36)
```

-   **Q16**. Calculate the mean *\"Percent of Population Fully Vaccinated\"* for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* \"2023-05-23\". Add this as a straight horizontal line to your plot from above with the `geom_hline()` function?

```{r}
vax.36_mean <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = T)
vax.36_mean 
```

```{r}
p <- ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated", title = "Vaccination Rate for La Jolla, CA 92037" ) +
  geom_hline(yintercept = vax.36_mean, linetype="dashed", col = "red")
p
```

-   **Q17.** What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the *\"Percent of Population Fully Vaccinated\"* values for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* \"2023-05-23\"?

    ```{r}
    summary(vax.36$persons_fully_vaccinated)
    ```

-   **Q18.** Using ggplot generate a histogram of this data.

```{r}
h <- ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(c(0,1))
  labs(x = "Percent Vaccinated", y="Count")
h
```

-   **Q19**. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

    ```{r}
    vax %>% filter(as_of_date == "2023-05-23") %>%  
      filter(zip_code_tabulation_area=="92040") %>%
      select(percent_of_population_fully_vaccinated)
    ```

```{=html}
<!-- -->
```
-   **Q20.** Finally make a time course plot of vaccination progress for all areas in the full dataset with a `age5_plus_population > 36144`.

    ```{r}
    vax.36.all <- filter(vax, age5_plus_population > 36144)


    ggplot(vax.36.all) +
      aes(as_of_date,
          percent_of_population_fully_vaccinated, 
          group=zip_code_tabulation_area) +
      geom_line(alpha=0.2, color="blue") +
      ylim(c(0,1)) +
      labs(x="Date", y="Percent Vaccinated",
           title="Vaccination Rate Across California",
           subtitle="Only areas with a population above 36k are shown") +
      geom_hline(yintercept = vax.36_mean, linetype="dashed")
    ```
